Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
This document summarizes the First Workshop on Sustainable Software
for Science: Practice and Experiences (WSSSPE1). The content is based
on a set of collaborative notes taken during the workshop, including
discussion about the keynote presentations, three themed panels that
discussed papers accepted at the workshop, other discussion,
cross-cutting issues, use cases, and conclusions. It joins a nascent
literature seeking to understand what drives software work in science
and how the reward systems of science function, thereby shaping the
types of software projects undertaken, including the extent to which
developers are motivated to build software for the long-term, for the
use of others, and whether to work collaboratively or separately.
Unique perspectives were captured about issues such as comprehensive
documentation, development and deployment practices; software
licenses; career paths for developers; identifiers for software
through a mechanism such as a Digital Object Identifier or publication
of a ``software paper''; and evidence of software contribution and
impact, recorded in online systems such as source code repositories
like GitHub. For truly sustainable software, there should be no
endpoint, as the software products continue to be used beyond a single
institution, grant, and developer or development team.

@@

Because progress in scientific research is dependent on the quality of
and accessibility to software at all levels, it is now critical to
address many challenges related to the development, deployment, and
maintenance of reusable software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists, researchers, and
students are able to learn and adopt software-related skills and
methodologies. Established researchers are already acquiring some of
these skills, and in particular a specialized class of software
developers is emerging in academic environments as an integral part of
successful research teams. This first WSSSPE workshop provided a forum
for discussion of the challenges around sustaining scientific
software, including contributed short papers in the form of both
positions and experience reports. These short papers, as well as notes
from the debates around them, have been archived to provide a basis
for continued discussion, and have fed into the collaborative writing
of this report.  Some of the workshop submissions have been extended
to full papers, which form part of the same special journal edition in
which this report appears. The workshop generated a high level of
interest; an estimated~90 to~150 participants were in attendance at
different times of the day. The interest and discussions have already
led to follow-up activities: A smaller Python-specific workshop is
planned to be held at the~2014 SciPy conference, and a follow-on
WSSSPE2 workshop has been proposed to, and meanwhile been accepted for
the SC14 conference. In addition, funds to support the workshops have
been obtained from the US National Science Foundation (NSF) and the
Gordon and Betty Moore Foundation, and the original workshop website
has been turned into a community website to engender further
discussion and progress. Additionally, a minisymposium at the~2014
Society for Industrial and Applied Mathematics (SIAM) Annual Meeting
on ``Reliable Computational Science'' (SIAM AN14,
\url{http://meetings.siam.org}) is being co-organized by a WSSSPE1
participant to further explore some of the key issues raised at the
workshop.

@@

>>>>

\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
This document summarizes the First Workshop on Sustainable Software
for Science: Practice and Experiences (WSSSPE1). The content is based
on a set of collaborative notes taken during the workshop, including
discussion about the keynote presentations, three themed panels that
discussed papers accepted at the workshop, other discussion,
cross-cutting issues, use cases, and conclusions. It joins a nascent
literature seeking to understand what drives software work in science
and how the reward systems of science function, thereby shaping the
types of software projects undertaken. These incentives can determine the extent to which
developers are motivated to build software for the long-term, for the
use of others, and whether to work collaboratively or separately.
Unique perspectives were captured about issues such as comprehensive
documentation, development and deployment practices, software
licenses and career paths for developers. Attribution systems that capture
evidence of software contribution and impact were also discussed. These include
mechanisms such as Digital Object Identifiers, publication
of ``software papers'' and the use of online systems, for example source code repositories
like GitHub. For truly sustainable software, there should be no
endpoint, as the software products continue to be used beyond a single
institution, grant, and developer or development team.

@@

Because progress in scientific research is dependent on the quality of
and accessibility to software at all levels, it is now critical to
address many challenges related to the development, deployment, and
maintenance of reusable software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists, researchers, and
students are able to learn and adopt software-related skills and
methodologies. Established researchers are already acquiring some of
these skills, and in particular a specialized class of software
developers is emerging in academic environments as an integral part of
successful research teams. This first WSSSPE workshop provided a forum
for discussion of the challenges around sustaining scientific
software, including contributed short papers in the form of both
positions and experience reports. These short papers, as well as notes
from the debates around them, have been archived to provide a basis
for continued discussion, and have fed into the collaborative writing
of this report.  Some of the workshop submissions have been extended
to full papers, which form part of the same special journal edition in
which this report appears. The workshop generated a high level of
interest; an estimated~90 to~150 participants were in attendance at
different times of the day. The interest and discussions have already
led to follow-up activities: A smaller Python-specific workshop is
planned to be held at the~2014 SciPy conference, and a follow-on
WSSSPE2 workshop has been accepted for
the SC14 conference. In addition, funds to support the workshops have
been obtained from the US National Science Foundation (NSF) and the
Gordon and Betty Moore Foundation, and the original workshop website
has been turned into a community website to engender further
discussion and progress. Additionally, a minisymposium at the~2014
Society for Industrial and Applied Mathematics (SIAM) Annual Meeting
on ``Reliable Computational Science'' (SIAM AN14,
\url{http://meetings.siam.org}) is being co-organized by a WSSSPE1
participant to further explore some of the key issues raised at the
workshop.

@@

Wilkins-Diehr was supported by the Extreme Science and Engineering Discovery Environment (XSEDE), 
NSF ACI-1053575.
